---
title: Finding and preparing spatial data
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
---

```{r}
#| label: load-r-libraries
#| message: false
library(sf)
library(tidyverse)
library(ggspatial)
library(rnaturalearth)
```

# Learning objectives

- Identify reliable sources for finding spatial data online
- Understand basic principles for evaluating data quality using metadata
- Make informed choices between common spatial data formats:

  - Shapefile
  - GeoJSON
  - GeoPackage

- Load vector spatial data from your own computer into R using `st_read()`
- Load simple tabular data using `read_csv()`

- Explain why working directories are crucial for loading files easily

- Perform basic data cleaning tasks on `sf` objects using `dplyr` functions

- Join tabular data to your spatial data based on a common identifier using `left_join`

- Save you cleaned spatial data to a file using `st_write()`

# Concepts: Finding your ingredients and checking quality

## Where to find spatial data: reliable sources

### Natural earth

- Global datasets at various scales
- Public domain 
- Might not have super-detailed local data

### OpenStreetMap

- A collaborative project
- Amazingly detailed, often the most up-to-date source for features like roads/buildings, global coverage, free to use
- Data quality varies by region

### GADM

- Database of Global ADministrative Areas
- Consistent source for administrative boundaries worldwide, differente levels of detail
- Boundaries might not perfectly match official national sources in every single case

### Government geospatial portals

In the case of Colombia:

- [Colombia en mapas](https://www.colombiaenmapas.gov.co/)

  - Iniciar sesion > Iniciar sesión con Google
  - VI > [Ver a CeM v1](https://www.colombiaenmapas.gov.co/ndexv1.html)

    - Filtrar > Entidad Territorial > Colombia
    - Temáticas > Límites > Departamentos de Colombia

      - Formato de descargar: GDB (`.gdb`), Geopackage (`.gpkg`), Shapefile

      - Use Geopackage (it is highly recommended)

#### Search strategy

- Need general world/country/state boundaries? Start with `rnaturalearth`
- Need detailed streets, buildings, cafes, parks? Start with `osmdata`
- need official local data? In the case of Colombia go to [Colombia en mapas](https://www.colombiaenmapas.gov.co/) 

### Is this data any good? Metadata and Quality Checks

- Source: 

  - Who created the data? 
  - Is the source trustworthy for your purpose

  **Answer**: For the case of Colombia was created by Instituto Geográfico Agustín Codazzi - IGAC, so it is trustworthy 

- Date

  - When was created?
  - When was it las t updated
  - It is recent enough for your map?

  **Answer**: It was created in `2025-12-30` and it will be updated in `2026-02-05`

- Description/Abstract

  - What does this data actually represent?
  - What do the columns in the attribute table mean?

- CRS

  - What coordinate reference system does it use?

- Scale/Resolution

  - At what level of detail was it created?

- Accuracy: 

  - Does the provider mention how accurate the locations or attributes are believed to be?

- Limitations/known issues: 

  - Are there any problems the creator is aware of?

- LIcense/Use Contraints

  - How are you allowed to use this data?
  - Do you need to five credit?
  - Can you use it for commercial purposes?

  **Answer**: Licencia: CC BY 4.0 

### Choosing Data Formats 

Simple rule: If available prefer GeoPackage `.gpkg`

### Ethical considerations: Data SElection & Bias

- Source Bias:

  - Whose data is it?
  - Does the source have a perspective?
  - Does it represent all grouos/areas fairly?

- Representation:

  - Does the dataset accurately represent the phenomenom?
  - Are certain categories oversimplified?
  Does it exclude important groups?

- Privacy

  - Does the data contain potentially sensitive location information about individuals?

# Hands-on: Loading and Cleaning Your Own data

## Import 

```{r}
#| message: false
#| results: hide
depart_col <- st_read(
  dsn = "../000_data/colombia/departamentos-de-colombia_2025-12-30.gpkg"
)
mun_col <- st_read(
  dsn = "../000_data/colombia/municipios-de-colombia-2025-12-30.gpkg"
)

world_boundaries_loaded <- st_read(
  dsn = "../000_data/mapping-worlds-with-r/chapter-3/world_boundaries.gpkg"
)
country_indicators_loaded <- read_csv(
  file = "../000_data/mapping-worlds-with-r/chapter-3/country_indicators.csv"
)
```

## CRS

```{r}
st_crs(depart_col)
```

## Visualization

```{r}
#| label: fig-depart-col
#| fig-cap: Colombian departments
depart_col |>
  ggplot() +
  geom_sf(
    aes(geometry = SHAPE)
  ) +
  annotation_scale(
    location = "bl"
  ) +
  annotation_north_arrow(
    location = "tr"
  ) +
  coord_sf(datum = st_crs(x = 4626)) +
  theme_minimal()
```

```{r}
#| label: fig-mun-col
#| fig-cap: Colombian municipalities
mun_col |>
  ggplot() +
  geom_sf(
    aes(geometry = SHAPE)
  )
```

## Basic data cleaning with `dplyr`

### Select and rename

```{r}
world_selected <- world_boundaries_loaded |>
  select(
    NAME_ENGL,
    ISO3_CODE,
    geom
  )

world_remaned <- world_selected |>
  rename(
    country_name = NAME_ENGL,
    iso3 = ISO3_CODE,
    geometry = geom
  )

world_ne <- ne_countries(
  scale = "medium",
  returnclass = "sf"
) |>
  select(
    adm0_a3,
    continent
  ) |>
  st_drop_geometry()
```

### Join

```{r}
world_remaned_with_cont <- world_remaned |>
  left_join(
    y = world_ne,
    by = join_by(iso3 == adm0_a3)
  )

world_data_joined <- world_remaned |>
  left_join(
    y = country_indicators_loaded,
    by = join_by(iso3 == iso3c)
  )
```

### Filter

```{r}
africa_only <- world_remaned_with_cont |>
  filter(continent == "Africa")
```

#### Check if `geometry` is empty

```{r}
africa_only |>
  filter(st_is_empty(geometry))

world_data_joined |>
  filter(st_is_empty(geometry))
```

#### Check if `geometry` is valid

```{r}
africa_only |>
  st_is_valid()

world_data_joined |>
  st_is_valid()
```

# Import

```{r}
st_read(
  dsn = "../000_data/mapping-worlds-with-r/chapter-3/world_data_cleaned_joined.gpkg"
) |>
  glimpse()
```