---
title: Your mapping cockpit
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
---

```{r}
#| label: load-libraries-r
library(sf)
library(rnaturalearth)
```

# Sneak peek: coordinate reference systems (CSR)

**The problem**: the earth is roundish, but our computer screens and maps are flat. A CRS is a set of rules defining:

- How the Earth's shape is modeled (the Datum, e.g., WGS84 usde by GPS)
- How locations from the curved model are projected onto a flat map (the projection, e.g., Mercator, UTM, Albers)

**Why it matters**: if 2 datasets use different rules (different CRSs), their coordinates mean differernt things, and the won't align geographically!

**The solution**: we need to make sure all data we want to combine or plot together uses the same CRS. We often transform data from one CRS to another.

**Checking CRS with `sf`**

The `sf` package makes this easy. When you load spatial data into an `sf` object, you can check its CRS rules.

# Import data

```{r}
world_sf <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)
```

# CRS

## Check CRS

```{r}
crs_info <- st_crs(world_sf)
crs_info
```

## Transform

```{r}
target_crs_robinson <- "ESRI:54030"
world_robinson_sf <- world_sf |>
  st_transform(crs = target_crs_robinson)
crs_info_robinson <- st_crs(world_robinson_sf)
crs_info_robinson
```

# Visualization

```{r}
par(mfrow = c(1, 2))

plot(
  st_geometry(world_sf),
  main = "Original (EPSG:4326)",
  key.pos = NULL, # Do not draw a legend
  reset = FALSE,
  border = "grey"
)
plot(
  st_geometry(world_robinson_sf),
  main = "Transformed (Robinson)",
  key.pos = NULL,
  reset = FALSE,
  border = "blue"
)

par(mfrow = c(1, 1))
```